---
title: "Untitled"
output: html_document
---

```{r}
library(ggplot2)
library(lubridate)
library(purrr)
library(dplyr)
library(naniar)

theme_set(theme_minimal())
```


```{r}
ems_data <- vroom::vroom(here::here("data/original/CFD_CARS_EMS_DATA_121616TO51320.csv"), col_types = paste0(rep("c", 78), collapse = ""))
```

Update the names to be more code friendly

```{r}
ems_data <- ems_data %>% 
  rename_with(~ tolower(gsub(r"([\. ])", "_", .x))) %>% 
  rename_with(~gsub(r"(\(.*)", r"()", .x)) %>%
  rename_with(~gsub(r"(_$)", "", .x)) %>% 
  mutate(incident_date = date(as.POSIXct(incident_date, tz = "GMT", format = "%m/%d/%Y %H:%M:%S")))

ems_data$incident_date[which(ems_data$incident_date == ymd("2000-01-01"))] <- NA # remove weird date from data

ems_data_no_dups <- ems_data %>% 
  mutate(num_na = rowSums(is.na(ems_data))) %>% 
  filter(!is.na(response_incident_number)) %>% 
  group_by(response_incident_number) %>% 
  top_n(1, num_na) # grabs the duplicate row with the most reported values
```



```{r}
min(ems_data$incident_date, na.rm = TRUE)
max(ems_data$incident_date, na.rm = TRUE)

```

```{r}
map_dbl(ems_data, ~sum(is.na(.))) / nrow(ems_data)
```

## Working with dates

```{r}
ems_data %>% 
  ggplot() +
  geom_histogram(aes(x = incident_date))
```

Lets find how many responses before 2016, and are they all just 2000

```{r}
sum(ems_data$incident_date < ymd("2016-01-01"), na.rm = TRUE)
```

There's one that's messing up the vis

```{r}
which(ems_data$incident_date == ymd("2000-01-01"))
```
```{r}
slice(ems_data, 176944)
```
 
Looks like its a real case, but I can't imagine the date is right, so I'm setting it to `NA`

```{r}
ems_data$incident_date[which(ems_data$incident_date == ymd("2000-01-01"))] <- NA
```

```{r}
ems_data %>% 
  ggplot() +
  geom_histogram(aes(x = incident_date), bins = 1269, boundary = ymd("2019-05-13")) +
  theme_minimal()
```

Plot incidents per day with no duplicates

```{r}
ems_data_no_dups %>% 
  ggplot() +
  geom_histogram(aes(x = incident_date), bins = 1269, boundary = ymd("2019-05-13")) +
  theme_minimal()
```

Explore weekly patterns in data:

```{r}
ems_data_no_dups %>% 
  mutate(incident_wday = wday(incident_date, label = TRUE)) %>% 
  ggplot(aes(x = incident_wday)) +
  geom_bar() +
  theme_minimal()
```



Find how many days are covered:

```{r}
max(ems_data$incident_date, na.rm = T) - min(ems_data$incident_date, na.rm = T)
```
 
Find the real first and last day
```{r}
min(ems_data$incident_date, na.rm = T)
max(ems_data$incident_date, na.rm = T)
```

Find days with most incidents

```{r}
ems_data_no_dups %>% 
  filter(!is.na(incident_date)) %>% 
  group_by(incident_date) %>% 
  count() %>% 
  arrange(desc(n))
```


## Visualizing missing data

```{r, fig.width=10, fig.height=8}
ems_data %>% 
  select(incident_type, 
         incident_date, 
         response_incident_number,
         response_ems_unit_call_sign,
         scene_gps_latitude,
         agency_name) %>% 
  gg_miss_upset()
```

Based on this, I'm going to use response_incident_number to find duplicates


Check the plot for no duplicates
```{r}
ems_data_no_dups %>% 
  select(incident_type, 
         incident_date, 
         response_incident_number,
         response_ems_unit_call_sign,
         scene_gps_latitude,
         agency_name) %>% 
  gg_miss_upset()
```

Look at number of NAs over dataset

```{r, fig.width= 9, fig.height= 7}
# As points
ems_data %>% 
  ggplot() +
  geom_point(aes(x = 1:nrow(ems_data), y = rowSums(is.na(ems_data)))) + 
  labs(y = "Number of NA's", x = "index") +
  ylim(0, 78)

# As density

ems_data %>% 
  ggplot() +
  geom_hex(aes(x = 1:nrow(ems_data), y = rowSums(is.na(ems_data))), color = "grey1", size = 0.2) + 
  labs(y = "Number of NA's", x = "index") +
  ylim(0, 78)
```

